---
title: "Spotify Analysis"
author: "Kevin"
date: "20 February 2021"
output: rmarkdown::github_document
---

```{r, warning=FALSE, echo=FALSE, include=FALSE}
library(tidyverse)
library(data.table)
library(lubridate)
library(sqldf)
library(knitr)
options(dplyr.summarise.inform = FALSE)
```

## Spotify Data 

```{r, echo=FALSE, warning=FALSE}
### Importing data with correct data types
spotify_raw <- fread('data/song_features.csv')
spotify_read <- spotify_raw %>% select(-V1) %>% 
  mutate(endTime=ymd_hm(endTime),
         album_release_date=ymd(album_release_date),
         completed=ifelse(msPlayed==duration_ms,1,0)) %>%
  filter(albumName!="")
attr(spotify_read$endTime, "tzone") <- "Asia/Kolkata"
```

```{r, echo=FALSE}
### Creating time features
spotify_time <- spotify_read %>% 
  mutate(year=year(endTime), month=month(endTime), day=day(endTime), week=week(endTime),
         date=date(endTime), wday=wday(endTime), hour=hour(endTime), minute=minute(endTime)) %>%
  arrange(endTime) %>%
  mutate(streamID = row_number())
```

```{r, echo=FALSE}
### Creating listening session id feature
listening <- spotify_time %>% 
  mutate(startTime = endTime - msPlayed*0.001, 
         lag_endTime = lag(endTime),
         timegap = as.integer((startTime) - lag(endTime)),
         endTime_neg1mins = endTime - minutes(1),
         endTime_10mins = endTime + minutes(10)
         ) %>%
  select(streamID, startTime, endTime, lag_endTime, endTime_10mins, endTime_neg1mins, msPlayed, timegap)
start_stream <- sqldf("select distinct(streamID)
      from ((select streamID, startTime from listening) a left join 
      (select streamID as streamID1, endTime_neg1mins, endTime_10mins from listening) b 
      on(a.startTime >= b.endTime_neg1mins and a.startTime <= b.endTime_10mins and a.streamID > b.streamID1))
      where streamID1 is null
      ")
start_stream <- start_stream %>% mutate(
  end_streamID = lead(streamID) - 1,
  session_id = row_number()
)
spotify <- sqldf("select a.*, b.session_id
      from spotify_time a left join start_stream b 
      on(a.streamID>=b.streamID and a.streamID<=b.end_streamID)")

last_session <- max(spotify$session_id, na.rm=TRUE) + 1
spotify <- spotify %>% replace_na(list(session_id = last_session))
kable(head(spotify))
```


## Most listened to songs 

```{r, echo=FALSE, warning=FALSE}
songs <- spotify %>% 
  group_by(artistName, trackName) %>%
  summarise(no_of_streams = n(), 
            total_minutes = sum(msPlayed)/60000,
            no_of_completed_streams = sum(completed)) %>%
  arrange(desc(no_of_streams)) %>%
  head(10) 
kable(songs)
```

### By number of streams
```{r, echo=FALSE}
songs %>% ggplot(aes(x=reorder(trackName, no_of_streams), y=no_of_streams)) + 
  geom_bar(stat="identity", aes(fill=trackName)) + coord_flip()
```

```{r, echo=FALSE, include=FALSE}
### By minutes
songs %>% ggplot(aes(x=reorder(trackName, total_minutes), y=total_minutes)) + 
  geom_bar(stat="identity", aes(fill=trackName)) + coord_flip()
```


```{r, echo=FALSE, include=FALSE}
### Longest songs
spotify %>% 
  filter(completed==1) %>%
  distinct(artistName, trackName, duration_ms) %>%
  mutate(runtime_mins = duration_ms/60000) %>%
  arrange(desc(runtime_mins)) %>% head(10) %>%
  ggplot(aes(x=reorder(trackName, runtime_mins), y=runtime_mins)) +
  geom_bar(stat="identity", aes(fill=trackName)) + coord_flip()
```

## Most listened to Artists

```{r, echo=FALSE}
### Most listened Artists by total number of songs played, distinct albums listened to, total duration listened to 
artists <- spotify %>%
  group_by(artistName) %>%
  summarise(songs_played=n(), distinct_songs=n_distinct(trackName),
            distinct_albums=n_distinct(albumName), total_duration_hours=sum(msPlayed)/3600000,
            completed_songs=sum(completed)) %>%
  arrange(desc(total_duration_hours)) %>% head(10)
kable(artists)
```

### By total duration
```{r, echo=FALSE}
artists %>% ggplot(aes(x=reorder(artistName, total_duration_hours), y=total_duration_hours)) + 
  geom_bar(stat="identity", aes(fill=artistName)) + coord_flip()
```


```{r, echo=FALSE, include=FALSE}
#### Artist with most runtime
artists %>% ggplot(aes(x=reorder(artistName, total_duration_hours), y=total_duration_hours)) + 
  geom_bar(stat="identity", aes(fill=artistName)) + coord_flip()
```


## Most listened to album

```{r, echo=FALSE, warning=FALSE}
albums <- spotify %>%
  group_by(artistName, albumName) %>%
  summarise(no_of_songs_played=n(),
            total_duration_hours=sum(msPlayed)/3600000,
            completed_songs=sum(completed)) %>%
  arrange(desc(no_of_songs_played)) %>% head(10)
kable(albums)

```

### By Total duration

```{r, echo=FALSE}
albums %>% ggplot(aes(x=reorder(albumName, total_duration_hours), y=total_duration_hours)) +
  geom_bar(stat="identity", aes(fill=albumName)) + coord_flip()
```


```{r, echo=FALSE, include=FALSE}
### Most listened to album by time
albums %>% ggplot(aes(x=reorder(albumName, total_duration_hours), y=total_duration_hours)) +
  geom_bar(stat="identity", aes(fill=albumName)) + coord_flip()
```

## Story of High Visceral Album

```{r, echo=FALSE}
spotify %>% 
  filter(albumName=='High Visceral, Pt. 1', year=='2020', week<25) %>%
  group_by(week,track_no) %>%
  summarise(count=n()) %>%
  arrange(week,track_no) %>%
  ggplot(aes(x=track_no,y=week)) + geom_tile(stat="identity", aes(fill=count)) + 
  scale_fill_gradient(low="sky blue", high="darkblue")
```

  It was a different album which i didn't like immediately, but after some weeks it grew on me. Initially, I liked the louder first half of the album, but later I started liking the mellower latter half of the album. Eventually, I moved on to some other album.

## My favorite artists through the months

```{r, echo=FALSE, warning=FALSE}
spotify %>% 
  filter(artistName %in% c('King Gizzard & The Lizard Wizard', 'A Beacon School', 'Psychedelic Porn Crumpets', 'Sufjan Stevens', 'Crumb'), year=='2020') %>%
  group_by(month,artistName) %>%
  summarise(count=n()) %>%
  arrange(month,artistName) %>%
  ggplot(aes(x=month,y=artistName)) + 
  geom_tile(stat="identity", aes(fill=count)) + 
  scale_fill_gradient(low="yellow", high="red")
```

  The shifting between artists is visible. There were quite a few KGATLW albums that I liked which kept me glued to their stuff till May till I got into PPC's High Visceral. There was one month there where I was really into Sufjan Steven's ambient album Aporia, a departure from my regular genres. After that, I wasn't into music a lot.
  
## Listening Sessions

```{r, echo=FALSE}
sessions <- spotify %>%
  mutate(starthour = hour(endTime - msPlayed*0.001)) %>%
  group_by(session_id) %>%
  summarise(no_of_songs = n(),
            total_runtime = sum(msPlayed)/1000,
            start_hour = head(starthour, 1)) %>%
  arrange(session_id)
kable(head(sessions))
```

There are about 720 listening sessions in about 1 year, so close to 2 sessions a day on average

## When do I listen most during the day?

```{r, echo=FALSE}
sessions %>% 
  group_by(start_hour) %>%
  summarise(no_of_sessions=n()) %>%
  ggplot(aes(x=start_hour, y=no_of_sessions)) +
  geom_bar(stat="identity", aes(fill=start_hour)) 
```


